---
title: "Model Evaluation"
author: "Gloria"
date: "9/6/2020"
output: 
    md_document:
        variant: markdown_github
    html_document: default  
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_dir = ".", output_format = "all") })      
---

## Summary 

The purpose is to take a brief look at data from [kaggle](https://www.kaggle.com/aungpyaeap/supermarket-sales), supermarket sales from 3 locations in Myanmar, and build a regression model to later be implemented ina shiny app. 

## Getting the Data

```{r, message=F}
library(dplyr)
library(lubridate)
library(car)
```

```{r, comment=''}
supermarket=read.csv('supermarket_sales.csv')

#supermarket=supermarket %>% subset(select=-c(gross.margin.percentage, gross.income)) %>% mutate(Time2=paste(Date, Time)) %>% mutate(Time2=parse_date_time(Time2, orders="m/d/Y H:M")) %>% subset(select=-c(Date, Time))

supermarket=supermarket %>% subset(select=-c(gross.margin.percentage, gross.income)) %>% mutate(Date2=paste(Date, Time)) %>% mutate(Date2=parse_date_time(Date2, orders="m/d/Y H:M")) %>% mutate(Time2=hm(Time)) %>% subset(select=-c(Date, Time))

str(supermarket)
```

## Feature Selection

1. Must select features that are independent of all others. Current features:

```{r, echo=F, comment=''}
names(supermarket)
```

Here, Total will be identified as the dependent variable ($y$), and the following will be considered as independent variables ($x$): Gender, City, Product.line, Customer.type, Unit.price, Quantity, Payment, Rating. 

2. Verify that there isn't linear dependence between the features:

```{r, comment=''}
fit=lm(Total ~ factor(Gender) + City + Product.line + Customer.type + Unit.price + Quantity + Payment + Rating, data=supermarket)

alias(fit)
```

No complete or partial dependence was report, therefor no linear dependence detected among the features. 

3. Examine each feature variability when correlated to the other features compared to when they are orthogonal: 

```{r, comment=''}
sqrt(vif(fit))[,1]
```

There isn't much variability in the results, but out of all of them, Product.line showed the greatest variability. 

## Model Selection 

Run ANOVA to determine which combination of features that contributes to the best model:

```{r, comment=''}
fit=lm(Total ~ Gender, supermarket)
fit1=update(fit, Total ~ Gender + City+ Customer.type + Rating)
fit2=update(fit1, Total ~ Gender + City + Customer.type + Rating + Unit.price + Quantity + Payment)
fit3=update(fit2, Total ~ Gender + City + Customer.type + Rating + Unit.price + Quantity + Payment + Time2)
anova(fit, fit1, fit2, fit3)
```

Based on the p-value for each nested model, fit2 model (Gender + City + Customer.type + Rating + Unit.price + Quantity + Payment) has the greatest impact on Total. 